---
title: "Untitled"
author: "Arun Kumar"
date: "2025-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Sys.setenv(XML_CONFIG = "/usr/bin/xml2-config")
#install.packages("XML", type = "source")
```

## ================================
## 1. Setup
## ================================

```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)

# Define base directory (single source of truth)
base_dir <- "/Volumes/NGS/tcga_geneexpression/Gene-Expression-Analysis"
data_dir <- file.path(base_dir, "GDCdata")

dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
```

## ================================
## 2. Explore project (optional)
## ================================

```{r}

gdcprojects <- getGDCprojects()
lihc_summary <- getProjectSummary("TCGA-LIHC")

print(lihc_summary)
```

## ================================
## 3. Build Query (STRICT + CLEAN)
## ================================

```{r}

query_TCGA <- GDCquery(
  project = "TCGA-LIHC",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",   # important (adds precision)
  experimental.strategy = "RNA-Seq",
  workflow.type = "STAR - Counts",
  access = "open",
  sample.type = c("Primary Tumor", "Solid Tissue Normal")
)

# Inspect query result
query_results <- getResults(query_TCGA)
cat("Total files in query:", nrow(query_results), "\n")
```

## ================================
## 4. Download (ROBUST)
## ================================

```{r}

GDCdownload(
  query = query_TCGA,
  directory = data_dir,
  method = "api",        # more stable than client
  files.per.chunk = 10
)
```
## ================================
## 5. Validate Download
## ================================

```{r}
files_downloaded <- list.files(data_dir, recursive = TRUE)

cat("Downloaded files:", length(files_downloaded), "\n")

# Optional strict check
if (length(files_downloaded) == 0) {
  stop("❌ No files detected. Download likely failed.")
}
```

## ================================
## 6. Prepare Data (CRITICAL STEP)
## ================================

```{r}

TCGA_data <- GDCprepare(
  query = query_TCGA,
  directory = data_dir,
  summarizedExperiment = TRUE
)

TCGA_data
```

## ================================
## 7. Extract Core Matrices
## ================================

```{r}
counts <- assay(TCGA_data)
meta   <- as.data.frame(colData(TCGA_data))

# Quick sanity check
dim(counts)
table(meta$sample_type)
```
## ================================
## 8. Save Clean Objects (FAST + SAFE)
## ================================

```{r}
saveRDS(TCGA_data, file = file.path(base_dir, "lihc_se.rds"), compress = FALSE)
saveRDS(counts,    file = file.path(base_dir, "lihc_counts.rds"), compress = FALSE)
saveRDS(meta,      file = file.path(base_dir, "lihc_metadata.rds"), compress = FALSE)
```

################################################################################
## ================================
## 9. Data Processing
## ================================
```{r}

library(SummarizedExperiment)
library(dplyr)
base_dir <- "/Volumes/NGS/tcga_geneexpression/Gene-Expression-Analysis"
TCGA_data <- readRDS(file.path(base_dir, "lihc_se.rds"))
counts <- assay(TCGA_data)
meta   <- as.data.frame(colData(TCGA_data))

## Quick checks
cat("Counts dim:", dim(counts), "\n")
cat("Metadata dim:", dim(meta), "\n")

table(meta$sample_type)
```

```{r}
## Check column names in metadata
colnames(meta)

## Check if rownames match
all(colnames(counts) == rownames(meta))
```

## ================================
## Gene Filtering (VERY IMPORTANT)
## Low-count genes create noise in
## PCA
## DESeq2 dispersion
## false DEGs
## ================================

```{r}
## Keep genes with sufficient counts
keep_genes <- rowSums(counts >= 10) >= 10
counts_filtered <- counts[keep_genes, ]

cat("Genes before filtering:", nrow(counts), "\n")
cat("Genes after filtering:", nrow(counts_filtered), "\n")
```

##=================================
## Basic sanity plots (quick check)
##=================================

```{r}
## Library size check
lib_sizes <- colSums(counts_filtered)
summary(lib_sizes)

## Detect extreme outliers
boxplot(log10(lib_sizes),
        main = "Library Size Distribution (log10)",
        ylab = "log10(read counts)")
```

```{r}
## Convert to factor with correct reference
meta$condition <- factor(meta$sample_type)
# Set NORMAL as reference (important for logFC direction)
meta$condition <- relevel(meta$condition, ref = "Solid Tissue Normal")

table(meta$condition)
```



```{r}
design_formula <- ~ condition

library(DESeq2)

dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered,
  colData   = meta,
  design    = design_formula
)

dds <- dds[rowSums(counts(dds)) > 1, ]
```


```{r}
dds <- DESeq(dds)
vsd <- vst(dds, blind = TRUE)
```

#======================
# PCA (Quality Control)
#======================

```{r}
library(ggplot2)

pca_data <- plotPCA(vsd, intgroup = "condition", 
                    returnData = TRUE)

percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal()
```

```{r}
ggplot(pca_data, aes(PC1, PC2, color = condition, shape = condition)) +
  geom_point(size = 3) +
  theme_minimal()
```

## outlier check
```{r}
# Identify extreme PCA outliers
pca_df <- as.data.frame(pca_data)

# Example: flag extreme PC1 values
outliers <- pca_df[abs(pca_df$PC1) > 80 | abs(pca_df$PC2) > 80, ]
outliers
```

##=================
## DEGs analysis
##=================
```{r}
res <- results(dds, 
               contrast = c("condition", 
                            "Primary Tumor", 
                            "Solid Tissue Normal"))
```
```{r}
resultsNames(dds)
```


```{r}
library(apeglm)

resLFC <- lfcShrink(
  dds,
  coef = "condition_Primary.Tumor_vs_Solid.Tissue.Normal",
  type = "apeglm"
)
resLFC
```


```{r}
res_df <- as.data.frame(resLFC)
res_df <- na.omit(res_df)
res_df <- res_df[order(res_df$padj), ]
```


```{r}
deg_sig <- res_df %>%
  dplyr::filter(padj < 0.05 & abs(log2FoldChange) > 1)
deg_sig
```

```{r}
deg_sig$ensembl <- gsub("\\..*", "", rownames(deg_sig))
deg_sig
```


```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)

gene_ids <- gsub("\\..*", "", rownames(deg_sig))

map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = gene_ids,
  keytype = "ENSEMBL",
  columns = c("SYMBOL", "GENENAME")
)
map$ENSEMBL <- gsub("\\..*", "", map$ENSEMBL)
map
```

```{r}
library(dplyr)

deg_annotated <- deg_sig %>%
  left_join(map, by = c("ensembl" = "ENSEMBL")) %>%
  filter(!is.na(SYMBOL)) %>%
  group_by(SYMBOL) %>%
  slice_max(abs(log2FoldChange), n = 1) %>%
  ungroup()
deg_annotated
```
















