---
title: "affyreadspnsp"
author: "Arun Kumar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r , echo=FALSE}
library(affy)
library(GEOquery)
library(tidyverse)
library(dplyr)
library(limma)
library(janitor)
#library(pheatmap)
```

```{r}
mouse_cd133_data <- ReadAffy(celfile.path = "data/CD133_Mouse_cellfiles/")
```

#performing RMA data normalization

```{r, echo=FALSE}

normalized.data <- rma(mouse_cd133_data)

```

## get expession values

```{r}
normalize_cd133mouse_expr <- as.data.frame(exprs(normalized.data))
```

## map affyprobe ids to gene symbol

```{r}
metadata <- read_csv("HT_MG-430_PM.na36.annot.csv", skip = 22, show_col_types = FALSE)
metadata
```


```{r}
metadata_MG430PM <- clean_names(metadata)
metadata_MG430PM
```

##subset meta data

```{r}
meta_sub <- metadata_MG430PM %>% 
  dplyr::select(c(probe_set_id,uni_gene_id,gene_title,gene_symbol,ensembl))

meta_sub 
write.csv(meta_sub,"meta_sub.csv")
```

## merge metadata with normalized expression values
## Normalize data

```{r}
normalize_cd133mouse_expr1 <- normalize_cd133mouse_expr %>% 
  rownames_to_column( var = "ID") %>% 
  inner_join(.,meta_sub, by = c("ID" = "probe_set_id"))

```

### final Normalize mouse data
```{r}
normalize_annotate_mouse_CD133 <- normalize_cd133mouse_expr1 %>% 
  select(c(48,49,1:47))
```

## DEGs analysis

```{r, echo=FALSE}
#eset <- rma(mouse_cd133_data)
pData(eset)
```

```{r}
# Sample names and condition (groups)


sample_names <- c(
  "K_CD133neg_F0064_18_726.ga.cel", "K_CD133neg_F0130_18_813.ga.cel", 
  "K_CD133neg_F0135_18_865.ga.cel", "K_CD133neg_F0138_18_853.ga.cel", 
  "K_CD133neg_F0152_18_809.ga.cel", "K_CD133neg_F0159_18_857.ga.cel", 
  "K_CD133neg_M0014_18_724.ga.cel", "K_CD133neg_M0099_18_734.ga.cel", 
  "K_CD133neg_M0132_18_728.ga.cel", "K_CD133neg_M0141_18_861.ga.cel", 
  "K_CD133pos_F0064_18_727.ga.cel", "K_CD133pos_F0130_18_814.ga.cel", 
  "K_CD133pos_F0135_18_866.ga.cel", "K_CD133pos_F0138_18_854.ga.cel", 
  "K_CD133pos_F0152_18_810.ga.cel", "K_CD133pos_F0159_18_858.ga.cel", 
  "K_CD133pos_M0014_18_725.ga.cel", "K_CD133pos_M0099_18_735.ga.cel", 
  "K_CD133pos_M0132_18_729.ga.cel", "K_CD133pos_M0141_18_862.ga.cel", 
  "K_WT_F0054_20_870.ga.cel", "K_WT_F0055_20_871.ga.cel", 
  "K_WT_M0030_20_872.ga.cel", "K_WT_M0075_20_869.ga.cel", 
  "T_CD133neg_F0064_18_718.ga.cel", "T_CD133neg_F0095WT_18_722.ga.cel", 
  "T_CD133neg_F0130_18_815.ga.cel", "T_CD133neg_F0135_18_867.ga.cel", 
  "T_CD133neg_F0138_18_855.ga.cel", "T_CD133neg_F0152_18_811.ga.cel", 
  "T_CD133neg_F0159_18_859.ga.cel", "T_CD133neg_M0014_18_716.ga.cel", 
  "T_CD133neg_M0132_18_730.ga.cel", "T_CD133neg_M0141_18_863.ga.cel", 
  "T_CD133pos_F0064_18_719.ga.cel", "T_CD133pos_F0095WT_18_723.ga.cel", 
  "T_CD133pos_F0130_18_816.ga.cel", "T_CD133pos_F0135_18_868.ga.cel", 
  "T_CD133pos_F0138_18_856.ga.cel", "T_CD133pos_F0152_18_812.ga.cel", 
  "T_CD133pos_F0159_18_860.ga.cel", "T_CD133pos_M0014_18_717.ga.cel", 
  "T_CD133pos_M0132_18_731.ga.cel", "T_CD133pos_M0141_18_864.ga.cel"
)

# Define the groups
groups <- c(rep("K_CD133neg", 10), rep("K_CD133pos", 10), rep("K_WT", 4), 
            rep("T_CD133neg", 10), rep("T_CD133pos", 10))

# Check sample names and corresponding groups
data.frame(Sample = sample_names, Group = groups)

```

```{r}
# Check the phenotype data
pData(eset)$group <- groups
head(pData(eset))

```

```{r}
# Assign groups to phenotype data
pData(eset)$group <- groups

# Ensure groups are treated as factors
pData(eset)$group <- factor(pData(eset)$group)

# Create the design matrix
design <- model.matrix(~ 0 + pData(eset)$group)

# Assign proper column names to the design matrix
colnames(design) <- levels(pData(eset)$group)

# Check the design matrix
head(design)


```


```{r}
# Fit the linear model using limma
fit <- lmFit(eset, design)

```

```{r}
# Perform empirical Bayes moderation
fit_ebayes <- eBayes(fit)
```

# Perform differential expression analysis for specific contrasts
# For example, comparing T_CD133pos vs T_CD133neg

```{r}
# Define the contrast for T_CD133pos vs T_CD133neg
contrast_T_cd133pos <- makeContrasts(T_CD133pos - T_CD133neg, levels = design)

# fit contrast
contrasts_fit_T_cd133pos <- contrasts.fit(fit_ebayes, contrast_T_cd133pos)
contrasts_fit_T_cd133pos_ebayes <- eBayes(contrasts_fit_T_cd133pos)

```

# For example, comparing K_CD133pos vs K_CD133neg
```{r}
# Define the contrast for T_CD133pos vs T_CD133neg
contrast_K_cd133pos <- makeContrasts(K_CD133pos - K_CD133neg, levels = design)

# fit contrast
contrasts_fit_K_cd133pos <- contrasts.fit(fit_ebayes, contrast_K_cd133pos)
contrasts_fit_K_cd133pos_ebayes <- eBayes(contrasts_fit_K_cd133pos)

```

# Extract differential expression results

```{r}
topTable_T_cd133pos <- topTable(contrasts_fit_T_cd133pos_ebayes,
                                number = Inf, 
                                adjust.method = "BH", 
                                sort.by = "B",
                                p.value = 1)

topTable_K_cd133pos <- topTable(contrasts_fit_K_cd133pos_ebayes,
                                number = Inf, 
                                adjust.method = "BH", 
                                sort.by = "B",
                                p.value = 1)

```



```{r}
topTable_T_cd133pos %>% 
  rownames_to_column( var = "ID") %>% 
  inner_join(.,meta_sub, by = c("ID" = "probe_set_id"))-> DEGS_T_cd133pos

topTable_K_cd133pos %>% 
  rownames_to_column( var = "ID") %>% 
  inner_join(.,meta_sub, by = c("ID" = "probe_set_id"))-> DEGS_K_cd133pos
```



```{r}
# Filter DEGs based on log fold  and pvalue 
Reg_DEGS_T_cd133pos <- subset(DEGS_T_cd133pos,
                               (logFC >= 0.6 | logFC <= -0.6) & (P.Value < 0.05))

Reg_DEGS_K_cd133pos <- subset(DEGS_K_cd133pos,
                               (logFC >= 0.6 | logFC <= -0.6) & (P.Value < 0.05))

write.csv(Reg_DEGS_T_cd133pos, "Reg_DEGS_T_cd133pos.csv")
write.csv(Reg_DEGS_K_cd133pos,"Reg_DEGS_K_cd133pos.csv")
  
```

## heatmap


```{r}
# Extract the row names from the 'ID' column of 'DEGs'
deg_genes <- Reg_DEGS_T_cd133pos$ID
# Filter the expression data to include only DEGs

deg_expression <- eset[match(deg_genes, rownames(eset)), ]

# Sort the DEGs based on significance or fold change values
sorted_deg_expression <- deg_expression[order(Reg_DEGS_T_cd133pos$logFC,
                                              decreasing = TRUE), ]

# Define the order of samples based on conditions or treatments
sample_order <- c("K_CD133neg_F0064_18_726.ga.cel", "K_CD133neg_F0130_18_813.ga.cel", 
  "K_CD133neg_F0135_18_865.ga.cel", "K_CD133neg_F0138_18_853.ga.cel", 
  "K_CD133neg_F0152_18_809.ga.cel", "K_CD133neg_F0159_18_857.ga.cel", 
  "K_CD133neg_M0014_18_724.ga.cel", "K_CD133neg_M0099_18_734.ga.cel", 
  "K_CD133neg_M0132_18_728.ga.cel", "K_CD133neg_M0141_18_861.ga.cel", 
  "K_CD133pos_F0064_18_727.ga.cel", "K_CD133pos_F0130_18_814.ga.cel", 
  "K_CD133pos_F0135_18_866.ga.cel", "K_CD133pos_F0138_18_854.ga.cel", 
  "K_CD133pos_F0152_18_810.ga.cel", "K_CD133pos_F0159_18_858.ga.cel", 
  "K_CD133pos_M0014_18_725.ga.cel", "K_CD133pos_M0099_18_735.ga.cel", 
  "K_CD133pos_M0132_18_729.ga.cel", "K_CD133pos_M0141_18_862.ga.cel", 
  "T_CD133neg_F0064_18_718.ga.cel", "T_CD133neg_F0095WT_18_722.ga.cel", 
  "T_CD133neg_F0130_18_815.ga.cel", "T_CD133neg_F0135_18_867.ga.cel", 
  "T_CD133neg_F0138_18_855.ga.cel", "T_CD133neg_F0152_18_811.ga.cel", 
  "T_CD133neg_F0159_18_859.ga.cel", "T_CD133neg_M0014_18_716.ga.cel", 
  "T_CD133neg_M0132_18_730.ga.cel", "T_CD133neg_M0141_18_863.ga.cel", 
  "T_CD133pos_F0064_18_719.ga.cel", "T_CD133pos_F0095WT_18_723.ga.cel", 
  "T_CD133pos_F0130_18_816.ga.cel", "T_CD133pos_F0135_18_868.ga.cel", 
  "T_CD133pos_F0138_18_856.ga.cel", "T_CD133pos_F0152_18_812.ga.cel", 
  "T_CD133pos_F0159_18_860.ga.cel", "T_CD133pos_M0014_18_717.ga.cel", 
  "T_CD133pos_M0132_18_731.ga.cel", "T_CD133pos_M0141_18_864.ga.cel",
  "K_WT_F0054_20_870.ga.cel", "K_WT_F0055_20_871.ga.cel", 
  "K_WT_M0030_20_872.ga.cel", "K_WT_M0075_20_869.ga.cel")


# Reorder the columns of 'sorted_deg_expression' based on 'sample_order'

sorted_deg_expression <- sorted_deg_expression[ ,sample_order]
# convert into matrix
sorted_deg_expression1 <- as.matrix(sorted_deg_expression)
```

########## test to add the gene names in row, but didn't work ####
```{r}
sorted_deg_expression <- as.data.frame(sorted_deg_expression)

sorted_deg_expression_col <- rownames_to_column(sorted_deg_expression, var = "probe_set_id")

sorted_deg_expression11 <- as.matrix(left_join(sorted_deg_expression_col,meta_sub, by = "probe_set_id") %>% 
  dplyr::select(gene_symbol,2:45))

sorted_deg_expression11
```
######

```{r}
write.csv(sorted_deg_expression1, "sorted_deg_expression1.csv")
```


#Create a heatmap using the pheatmap package
```{r, echo=FALSE, fig.height= 8, fig.width=8}
library(pheatmap)

pheatmap(sorted_deg_expression1, 
         scale = "row", 
         clustering_distance_rows = "euclidean", 
         show_rownames = FALSE,
         col = colorRampPalette(c("blue", "white", "red"))(100))
```



```{r, echo=FALSE, fig.height= 10, fig.width=8}
# Calculate the distance matrix for clustering
dist_matrix <- dist(t(sorted_deg_expression1))

# Cluster the rows (genes) using hierarchical clustering
hc_rows <- hclust(dist_matrix)

# Reorder the rows of 'sorted_deg_expression' based on clustering
sorted_deg_expression <- sorted_deg_expression1[hc_rows$order, ]

# Create a heatmap with correct sample grouping
heatmap(sorted_deg_expression1, scale = "row", col = colorRampPalette(c("blue", "white", "red"))(100))
```


###############

```{r, echo=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
gene_list <- read.csv("spfa_genelist.csv")
# Convert gene symbols to Entrez IDs
gene_ids <- bitr(gene_list$Gene.Symbol, 
                 fromType = "SYMBOL", 
                 toType = "ENTREZID", 
                 OrgDb = org.Hs.eg.db)


```


```{r}
z_scores <- scale(sorted_deg_expression1)

```

##Heatmap for Epcam+FA


```{r}
CD133dmso_all <- read_delim("affydata/heatmap_zscore/CD133P DMSO (SP) CD133N DMSO (NSP).txt", "\t", show_col_types = FALSE)

CD133FA_all <- read_delim("affydata/heatmap_zscore/CD133P FA (SP) CD133N FA (NSP).txt", "\t", show_col_types = FALSE)
```


```{r}
CD133all <- inner_join(CD133dmso_all,CD133FA_all, by = "ID")
cd133all1 <- CD133all
```


```{r}
rownames(cd133all1) <-cd133all1$ID

```

```{r}
# Convert the tibble to a data frame
my_cd133all1 <- as.data.frame(cd133all1)

```



```{r}
# Rename the column header from "Probe Set ID" to "ID"
colnames(meta.sub)[colnames(meta.sub) == "Probe Set ID"] <- "ID"

# Check the updated column names
colnames(meta.sub)

```

```{r}
my1_cd133all1 <- inner_join( meta.sub,my_cd133all1, by = "ID")

```

```{r}

CD133PFADEG <- read.delim("affydata/heatmap_zscore/CD133PFA_DEGs_GeneSymbol.txt")
colnames(CD133PFADEG)[colnames(CD133PFADEG) == "Gene.Symbol"] <- "Gene Symbol"
CD133PFADEG <- distinct(CD133PFADEG)
```

```{r}
my_cd133PFADEG1 <- inner_join(my1_cd133all1,CD133PFADEG, by = "Gene Symbol")

```

```{r}
my_cd133PFADEG2 <- my_cd133PFADEG1 %>% dplyr::select(-ID) 

#rownames(my_cd133PFADEG2) <-my_cd133PFADEG2$`Gene Symbol`

```


```{r}
my_cd133PFADEG2 <- as.data.frame(my_cd133PFADEG2)
```

```{r}
my_cd133PFADEG2 <- distinct(my_cd133PFADEG2)
```

```{r}
sum(is.na(my_cd133PFADEG2))

```


```{r}

```



















